---
title:  "**Term Project Part 1 - Police shooting data (v2)**"
output:
  pdf_document: default
  word_document: default
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

```{r code0}
# Use the police shooting data and what you've learned so far in this course 
# to answer the following questions:
# 1. How many male and female victims were presented in the data set?
# 2. What are the characteristics of male vs. female victims in terms 
# of age, race, weapons, etc? 
# Using race as an example, how many male victims were White, Black,
# Hispanic, Native American, Asian, and Other,
# in comparison, how many female victims were White, Black, 
# Hispanic, Native American, Asian, and Other?  
# 3. How would you summarize your findings? Post your code and
# write a paragraph. These, and all future exercises, 
# will also be part of your project report. 

rm(list = ls())

library(ggplot2)
library(reshape2)
library(lubridate)
library(dplyr)
library(gridExtra)
library(tidyverse)
library(hrbrthemes)
library(ggthemes)

# Load the data
shooting_orig <- read_csv("fatal-police-shootings-data1.csv", 
                          col_names = TRUE, na="",  show_col_types = FALSE)
shooting_orig <- na.omit(shooting_orig)

cbPalette <- c("#F0E442", "#E69F00", "#56B4E9", "#009E73","#999999" ,
               "#0072B2", "#D55E00", "#CC79A7","#BAAAA3")
custom_color = c("darkgreen", "red","blue") 


#How many male and female victims were presented in the data set?
# sort(table(shooting_orig$gender), decreasing =FALSE)
count(shooting_orig, gender, sort = TRUE, name="Male/Female victims Count")


# How many male & female victims were presented in the data set.
shdat_gender <- shooting_orig %>%
  filter(!is.na(gender))%>%
  group_by(gender) %>%
  arrange(gender)%>%
  summarise(count = n())

shdat_gender1 <- shooting_orig %>% 
  group_by(gender) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))

# race vs gender
shdat_race <- shooting_orig %>%
  filter(!is.na(gender))%>%
  filter(!is.na(race))%>%
  group_by(gender, race) %>%
  summarise(count = n(), .groups = "drop")

shdat_race1 <- shooting_orig %>%
  filter(!is.na(gender))%>%
  filter(!is.na(race))%>%
  group_by(gender, race, age) %>%
  summarise(count = n(), .groups = "drop")


# Male vs Race
shdat_male <- shooting_orig %>%
  filter(!is.na(gender))%>%
  filter(!is.na(race))%>%
  filter(gender=="male")%>%
  group_by(gender, race) %>%
  tally() 

# Female vs Race
shdat_female <- shooting_orig %>%
  filter(!is.na(gender))%>%
  filter(!is.na(race))%>%
  filter(gender=='female')%>%
  group_by(gender, race) %>%
  tally() 

# Create group armedType for weapons 
count(shooting_orig, armed_with, sort=TRUE)
# unique(shooting_orig$armed_with)

group <- function (string){
  if(is.na(string)) return ("NA")
  if(string == "unarmed") return ("Unarmed")
  else if (string %in% c("undetermined", "unknown")) return ("Undetermined")
  else if (string == "vehicle") return ("Vehicle")
  else if (string %in% c("gun", "gun;knife", 
                         "other;gun",  "replica",
                         "gun;vehicle", "vehicle;gun"))  return ("Gun")
  else if (string %in% c("knife", "blunt_object", "blunt_object;blunt_object",
                         "sharp object", "knife;blunt_object")) 
                      return ("Sharp Object")
  else return ("Other")
}
#apply group on each element of a vector
shooting_orig$armedType <- sapply(shooting_orig$armed_with, group)

# create age category to consolidate different age groups
# Identify different characteristics based on age, weapons, race and gender
shdata_char <- shooting_orig %>%
  mutate(ageCategory = case_when(between(age, 0, 10) ~ "0 - 10",
                                 between(age, 11, 20) ~ "11 - 20",
                                 between(age, 21, 30) ~ "21 - 30",
                                 between(age, 31, 40) ~ "31 - 40",
                                 between(age, 41, 50) ~ "41 - 50",
                                 between(age, 51, 60) ~ "51 - 60",
                                 between(age, 61, 70) ~ "61 - 70",
                                 between(age, 71, 80) ~ "71 - 80",
                                 between(age, 81, 91) ~ "81 - 91")) %>%
  select(gender, age, race, ageCategory, armedType, )

# Male victims age & armed type chars
shdata_male_chars <- shdata_char %>% 
  filter(gender=='male')%>%
  group_by(ageCategory) %>% 
  select(ageCategory, gender, race, armedType)

# female victims age & armed type chars
shdata_female_chars <- shdata_char %>% 
  filter(gender=='female')%>%
  group_by(ageCategory) %>% 
  select(ageCategory, gender, race, armedType)

```
```{r code1}
# 1. How many male and female victims were presented in the data set?

# bar plots to show gender counts in the dataset

g1 <- ggplot(shdat_gender1, aes(x = "", y = perc, fill = gender)) +
  geom_col(color = "black") +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5)) +
  ggtitle("Male/Female victims percentage")+
  guides(fill = guide_legend(title = "Gender")) +
  # scale_fill_viridis_d() +
  scale_fill_manual(values=cbPalette)+
  coord_polar(theta = "y")  

g2 <- ggplot(shdat_gender, aes(x=gender, y=count, fill=gender)) + 
  geom_bar(stat = "identity", width=0.2) +
  # scale_fill_manual(values=cbPalette)+
  scale_fill_colorblind()+
  ggtitle("Male/Female victims count")+
  labs(y = "Count", x = "Gender" )+
  scale_y_continuous(breaks = seq(0,2800,200), limits = c(0, 2800) )+
  theme_minimal()+
  annotate(geom = 'text', label = 'A ', x = Inf, y = -Inf, 
           hjust = 1, vjust = -1, size =10)+
  # coord_flip()+
  theme(text = element_text(size = 11))


grid.arrange(g2,g1, ncol=2)
# grid.arrange(g3, ncol=1)

```

## Figure 1. Panel A shows Male/Female victims count in the datatset using Bar chart and 
## Pie chart shows percentage distribution between male & female victims.

\newpage

```{r code2}

# 2. Using race as an example, how many male victims were White, Black,
# Hispanic, Native American, Asian, and Other,
#  in comparison, how many female victims were White, Black, 
# Hispanic, Native American, Asian, and Other? 

g3 <- ggplot(shdat_race, aes(x=gender, y= count, fill=race)) +
  geom_bar(stat="identity", position=position_dodge())+
  scale_y_continuous(breaks = seq(0,1500,300), limits = c(0, 1500))+
  scale_fill_colorblind()+
  annotate(geom = 'text', label = 'A ', x = Inf, y = -Inf, 
           hjust = 1, vjust = -1, size =12)+
  labs(title = "Race distribution by Gender",
       x = "Gender", y="Count") +
  theme(axis.text.x = element_text(size=11, angle=0), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))+
 theme(text = element_text(size = 11))+
 theme_clean()
 # hjust = 1, vjust = -1, size =12)+
 # theme(text = element_text(size = 12, family = 'Garamond'),
 #       legend.position = 'bottom',
 #       legend.direction = 'horizontal')


#Race and Gender Plot
g4 <- ggplot(shooting_orig[!(shooting_orig$race %in% c(NA)), ],
              aes(x=race, fill=gender)) + geom_bar() +
  scale_y_continuous(breaks = seq(0,1500,250), limits = c(0, 1500))+
  facet_wrap(~gender) + 
  labs(title = "Shooting Distribution by Race and Gender",
       x = "Race", y="Count") +
  geom_text( aes(label=after_stat(count)),
             stat='count', nudge_x=-0.14, nudge_y=0.125,width=2 )+
  annotate(geom = 'text', label = 'B ', x = Inf, y = -Inf, 
           hjust = .8, vjust = -1, size =12)+
  scale_fill_manual(values=cbPalette)
# scale_fill_colorblind()


grid.arrange(g3, ncol=1)
grid.arrange(g4, ncol=1)
# # grid.arrange(g9, ncol=1)
# grid.arrange(g10, ncol=1)
# grid.arrange(g11, ncol=1)

```

## Figure 2. Panel A shows Race distribution by gender. 
## panel B shows Shooting distribuition by age & gender.

\newpage

```{r code3}
# # 3. What are the characteristics of male vs. female victims 
# in terms of age, race, weapons, etc? 

g5 <- ggplot(shdata_female_chars, aes(x=race, y=ageCategory , fill=armedType)) +
  geom_bar(stat="identity", position=position_dodge())+
  # ggtitle("Female victims race, age and armed type characteristics")
  # scale_fill_manual(values=cbPalette)+
  scale_fill_colorblind()+
  labs(title = "Female victims characteristics by race, age and armed type",
       x = "Race", y="Age Category in Years") +
  annotate(geom = 'text', label = 'A ', x = Inf, y = -Inf, 
           hjust = 0.6, vjust = -1, size =11)+
  theme(axis.text.x = element_text(size=11, angle=0), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))+
  theme(text = element_text(size = 11))+
  theme_clean()

g6 <- ggplot(shdata_male_chars, aes(x=race, y=ageCategory , fill=armedType)) +
  geom_bar(stat="identity", position=position_dodge())+
  # geom_bar(stat = "identity", width = 0.5) +
  ggtitle("Male victims race, age and armed type characteristics")+
  scale_fill_colorblind()+
  labs(title = "Male victims characteristics by race, age and armed type",
       x = "Race", y="Age Category in Years") +
  annotate(geom = 'text', label = 'B ', x = Inf, y = -Inf, 
           hjust = 0.6, vjust = -1, size =12)+
  theme(axis.text.x = element_text(size=11, angle=0), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))+
  theme(text = element_text(size = 11))+
  theme_clean()


# Weapon distribution by Race and Gender Plot
g7 <- ggplot(shooting_orig,
              aes(x=race, fill=armedType)) + geom_bar() +
  scale_y_continuous(breaks = seq(0,1450,100), limits = c(0, 1450))+
  facet_wrap(~gender) + 
  labs(title = "Weapon distribuition by Race and Gender",
       x = "Race", y="Count") +
  # geom_text( aes(label=after_stat(count)),
  #            stat='count', nudge_x=-0.14, nudge_y=0.125,width=2 )+
  annotate(geom = 'text', label = 'C ', x = Inf, y = -Inf, 
           hjust = 0.6, vjust = -1, size =11)+
  theme_clean()
  # scale_fill_manual(values=cbPalette)
  # scale_fill_colorblind()

grid.arrange(g5, ncol=1)
grid.arrange(g6, ncol=1)
grid.arrange(g7, ncol=1)

```

## Figure 3. Panel A shows Female victims characteristics based on age, race, weapons.
## Panel B shows Male victims characteristics based on age, race, weapons.
## Panel C shows Weapon distribuition by Race and gender.

\newpage

```{r code4}
g8 <- ggplot(shooting_orig, aes(x=state, fill=race)) +
  geom_bar() +
  ggtitle("Shooting distribution by States") +
  # scale_y_continuous(expand = c(0, 0)) +
  labs(y = "Count", x = "States" )+
  scale_y_continuous(breaks = seq(0,450,50), limits = c(0, 450))+
  # scale_fill_manual(values=cbPalette)+
  annotate(geom = 'text', label = 'A ', x = Inf, y = -Inf, 
           hjust = 0.6, vjust = -2, size =10)+
  theme(axis.text.x = element_text(size=8, angle=90), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))+
  theme(text = element_text(size = 11))
grid.arrange(g8, ncol=1)

```

## Figure 4. Extra Plot - Panel A gives an overall view of how the shooting distribuition by state wise looks for various races.

\newpage

**Summary:**

### Figure1:
Police shooting data set csv file was loaded to the data frame and analyzed. Total number of male and
female victims count from the given data set are shown. Overall there are more number male victims than
female victims(about 95% are males and only 5% are females)
Male - 2711, Female - 137

### Figure2:
When looked at the race distribution of male/female victims, White race tops both the male/female lists.
Next on the list is black race and the least one is Others.

### Figure3:
When different characteristics like age, race, gender and weapons are analyzed, 
male victims age ranges from 6 years and up to 91 years and female victims age ranges from 11 to 80 years.
Both male and female victims use and carry variety of armed types 
except in female there is no "Other" type.
Most races of female victims limit their weapons to gun, other and sharp objects. 
Only black & white races of females have older people, whereas in male victims 
older peoples are in almost all the races









**Team members contribution:**

Sivarajan Jaganathan & Ripa Shah:
Table count
Figure 1: Bar Chart/Pie Chart
Figure 4: Panel A
Summary


Sivarajan Jaganathan
Figure 2: Panel A 
Figure 3: Paenl A & B

Ripa Shah
Figure 2: Panel B 
Figure 3: Panel C



```{r code0, eval=FALSE, echo=FALSE}
```
```{r code1, eval=FALSE, echo=FALSE}
```
```{r code2, eval=FALSE, echo=FALSE}
```
```{r code3, eval=FALSE, echo=FALSE}
```
```{r code4, eval=FALSE, echo=FALSE}
```
